PAGAI : a path sensitive static analyser

### 2012-05-19 papers/TAPAS2012/pagai.tex
bla bla sur les variables

LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.

SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge to this control node was last taken. For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$. In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.

This motivates a key implement decision of our tool: only those variables
$v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as
coordinates in the abstract domain (e.g. as dimensions in polyhedra), assuming
they are live at the associate control point.

	For instance, assume that $x,y,z$ are numerical variables of a program,
	$x$ is defined as $x = y+z$, and $x,y,z$ are live at point $p$. Instead of having
	$x$ as a dimension for the abstract value at point $p$, we only have $y$ and $z$. All the properties
	for $x$ can be directly extracted from the abstract value attached to $p$ and the information $x=y+z$.
	This is an optimisation in the sense that there is redundant information in
	the abstract value if both $x,y$ and $z$ are dimensions of $X_p$.

@@

A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON. This has three benefits:
(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.

>>>>

LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
Because LLVM generally assigns rather straightforward names (e.g. \lstinline|x.0| for the first renaming of variable \lstinline|x|), the user can map the invariants back to the original source code; an automatic back-to-source mapping, based on debugging information, is being developed.

LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.

SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge was last taken to reach this control node. For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$.

In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as arithmetic functions, because they are defined using $\phi$-functions, loads from memory, return values from function calls, or other numerical operations (e.g. bitwise operators) that are not representable with our class of basic arithmetic operations. We may even vary the class of arithmetic operations, for instance, by restricting ourselves to linear ones.

This motivates a key implement decision of our tool: only those variables
$v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as
coordinates in the abstract domain (e.g. as dimensions in polyhedra), assuming
they are live at the associated control point.

	For instance, assume that $x,y,z$ are numerical variables of a program,
	$x$ is defined as $x = y+z$, and $x,y,z$ are live at point $p$. Instead of having
	$x$ as a dimension for the abstract value at point $p$, we only have $y$ and $z$. All the properties
	for $x$ can be directly extracted from the abstract value attached to $p$ and the relation $x=y+z$.
	This is an optimisation in the sense that there is redundant information in
	the abstract value if both $x,y$ and $z$ are dimensions of $X_p$.

@@

A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON. This has three benefits:
(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;%
\footnote{The additional dimensions express linear equalities between variables, which could be dealt with efficiently within polyhedra libraries, which anyway keep a basis of the linear affine variety containing the polyhedra.
They should therefore cost little assuming some sparse representation of the constraints.
Alas, several libraries, including APRON, compute with \emph{dense} vectors and matrices, which means that any increase in dimensions slows computations.}
(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.
